F1000 paper figures

### 2017-06-09 inst/paper/figs.Rmd
update figs and add tutorial

---
title: "F1000 paper figures"
author: "Kent Riemondy RBI"
date: "June 6, 2017"
output: html_document
---

@@

pub_theme <- theme(axis.text = element_text(size = 14),
                   plot.title = element_text(size = 14),
                   strip.text = element_text(size = 14))

@@

```{bash}
# Function to benchmark any bash command
repeats=10

@@

mean_dat <- group_by(dat, expr, package) %>% 
             summarize(mean = mean(time), se = sd(time))

mean_dat$func <- ifelse(mean_dat$package == "BEDTools", 
                        mean_dat$expr, 
                        str_split(mean_dat$expr, "\\(",  simplify = T)[, 1] %>% 
  str_split(., "_", simplify = T) %>% 
  .[,2])

p <- ggplot(mean_dat, aes(x=reorder(func, mean), y = mean, fill = package)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), 
                width = .1, position = position_dodge(1)) +
  scale_fill_grey() + 
  coord_flip() +
  theme_bw() +
  labs(
    y='execution time (seconds)',
    x='',
    subtitle = paste(comma(n), "random 1 kilobase intervals,", 
                     comma(nrep), "repetitions")) +
  theme_classic() +
  pub_theme +
  theme(
    legend.text = element_text(size = 14),
    legend.title = element_blank(),
    plot.subtitle = element_text(size = 10)
  )

plot_grid(p, align = 'h', nrow = 1, labels="AUTO")
ggsave("figure3.pdf", width = 5.6, height = 4)

@@

>>>>

---
title: "F1000 paper figures"
author: "Kent Riemondy RBI"
date: "June 6, 2017"
output: pdf_document
---

@@

pub_theme <- theme(axis.text = element_text(size = 20),
                   axis.title = element_text(size = 20),
                   plot.title = element_text(size = 20),
                   strip.text = element_text(size = 20))

@@

```{bash, eval = F}
# Function to benchmark any bash command
repeats=10

@@

#mean_dat <- group_by(dat, expr, package) %>% 
#             summarize(mean = mean(time), se = sd(time))

dat$func <- ifelse(dat$package == "BEDTools", 
                   dat$expr, 
                   str_split(dat$expr, "\\(",  simplify = T)[, 1] %>% 
  str_split(., "_", simplify = T) %>% 
  .[,2])

# remove makewindows, really throws off proportion of graph
dat <- filter(dat, func != "makewindows")
valr_data <- filter(dat, package == "valr")

plot_a <- ggplot(res, aes(x=reorder(expr, time), y = time)) +
  geom_boxplot(width = 0.75, size = 0.33, outlier.shape = NA, fill = RColorBrewer::brewer.pal(3, "Set1")[2]) +
  coord_flip() +
  theme_bw() +
  labs(
    y='execution time (seconds)',
    x='',
    subtitle=paste0(comma(n), " random x/y intervals,\n", comma(nrep), " repetitions")) +
  theme_classic() +
  pub_theme +
  theme(
    legend.text = element_text(size = 14),
    legend.title = element_blank(),
    legend.position = "none",
    plot.subtitle = element_text(size = 18),
    axis.text.x = element_text(angle = 90)
  )

plot_b <- ggplot(dat, aes(x=reorder(func, time), y = time, fill = package)) +
  geom_boxplot(width = 0.75, size = 0.33, outlier.shape = NA) +
  scale_fill_brewer(palette = "Set1") + 
  coord_flip() +
  theme_bw() +
  labs(
    y='execution time (seconds)',
    x='') +
  theme_classic() +
  pub_theme +
  theme(
    legend.text = element_text(size = 18),
    legend.title = element_blank(),
    legend.position = "top",
    axis.text.x = element_text(angle = 90)
  )

plot_grid(plot_a, plot_b, align = 'h', nrow = 1, labels="AUTO", label_size = 28)
ggsave("figure3.pdf", width = 20, height = 10)

@@

## Code demonstrations

```{r}
library(valr)
# function to retrieve path to example data
bed_filepath <- valr_example("3fields.bed.gz") 
read_bed(bed_filepath)

#using URL
read_bed("https://github.com/rnabioco/valr/raw/master/inst/extdata/3fields.bed.gz")
```

Using mysql
```{r, warnings = F}
library(dplyr, warn.conflicts = F)
hg19 <- db_ucsc("hg19")
tbl(hg19, "refGene")
```

### Interval statistics

Estimates of significance for interval overlaps can be obtained by combining `bed_shuffle()`, `bed_random()` and the `sample_` functions from `dplyr` with interval statistics in `valr`.

Here we examine the overlap of repeat classes in the human genome (on `chr22` only, for simplicity) using `bed_jaccard()`.

```{r reldist_shuffle, message = FALSE, warning = FALSE}

repeats <- read_bed(valr_example('hg19.rmsk.chr22.bed.gz'), n_fields = 6) 
genome <- read_genome(valr_example('hg19.chrom.sizes.gz'))
genes <- read_bed12(valr_example('hg19.refGene.chr22.bed.gz'))
exons <- bed12_to_exons(genes)

shuffle_intervals <- function(n, .data, genome) {
  replicate(n, bed_shuffle(.data, genome), simplify = FALSE) %>%
    bind_rows(.id = 'rep') %>%
    group_by(rep) %>% nest()
}

shuffled <- shuffle_intervals(n = 100, repeats, genome) %>%
  mutate(jaccard = data %>%
           map(bed_jaccard, repeats) %>%
           map_dbl("jaccard"))

shuffled

bed_jaccard(repeats, exons)
```

## Correlations among DNase I hypersensitive sites

Here we use `bed_jaccard` for a large-scale comparison of related datasets. As shown in the [`BEDtools` tutorial](http://quinlanlab.org/tutorials/bedtools/bedtools.html), we can measure the similarity of DNaseI hypersensitivity sites for 20 fetal tissue samples.

This data was taken from [Maurano *et al.* Systematic Localization of Common Disease-Associated Variation in Regulatory DNA. (2012) *Science*](www.sciencemag.org/content/337/6099/1190.short).

```{r init, warning = FALSE, message = FALSE}
library(valr)
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(stringr)
library(ggplot2)
library(ggrepel)
library(ComplexHeatmap)
```

First read all 20 BED files containing DNase I hypersensitivity sites from 20 fetal tissues.

```{r read_files, eval = F}
dnase_files <- list.files('data/dnasei',pattern = 'merge.bed.gz', full.names = TRUE)
data <- dnase_files %>% map(read_bed, n_fields = 4)
```

Then generate a 20x20 table containing a Jaccard statistic for each of the 400 pairwise comparisons.

```{r jaccard-mat, eval = F}
res <- data %>%
  cross2(.,.) %>%
  map(lift(bed_jaccard)) %>% 
  map("jaccard") %>%
  flatten_dbl() %>%
  matrix(nrow = 20, ncol = 20)
```

We also need to generate labels for the table from the file names.

```{r jaccard-names, eval = F}
# names are tissue + sample_num
col_names <- dnase_files %>%
  str_split('/') %>% map(`[[`, 3) %>%
  str_split('\\.') %>% map(`[[`, 1) %>%
  str_split('-') %>% map(`[[`, 1) %>%
  flatten_chr() %>%
  str_replace('^f', '') %>%
  str_c(str_c('-', seq(length(.))))

colnames(res) <- col_names
rownames(res) <- col_names
```

Now the Jaccard coefficients can be visualized in heatmap form.

```{r jaccard-heatmap, fig.cap='Heatmap of Jaccard scores', fig.align='center', fig.width=8, eval = F}
Heatmap(res, color_space = 'Blues')
```

Finally we can do some PCA analysis on the Jaccard coefficients to identify clusters of related cell types.

```{r jaccard-PCA, fig.cap='PCA analysis of Jaccard scores', fig.align='center', fig.width=8, eval = F}
pca <- broom::tidy(prcomp(res)) %>% as_data_frame()

pca_comps <- filter(pca, PC <= 2) %>%
  tidyr::spread(PC, value) %>%
  setNames(c('label','PC1','PC2'))

ggplot(pca_comps, aes(PC1, PC2)) +
  geom_point(size = 3, color = 'red') +
  geom_text_repel(aes(label = label))
```
